Supermartingales and Termination Analysis of Statistical PCF

### 2020-06-05 probabilisticProgrammingMartingales.tex
More examples

\paragraph{}
In both of these cases, there are only some values of the ranking function that are semantically important. Define a \emph{partial ranking function} on a closed term $M$ to be a partial function 
$f : Rch(M) \rightharpoonup \mathbb R$ 
%$f : Rch(M) \nrightarrow \mathbb R$ 
such that
%\lo{NOTATION. Common notation for partial function $\rightharpoonup$}
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability. Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
A partial ranking function that is total is just a ranking function. Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.

@@

\paragraph{Results to prove:}
\begin{enumerate}
\item (AST Soundness) If a closed SPCF term $M$ is antitone-convex rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$. \lo{This should be fine.}

@@

For a more complex example, consider the following ``continuous random walk''
\[
\underbrace{\big
(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample) \oplus_{1/2} f(x - \tsample)} \big)}_{\Xi} \, 10
\]
Let $g(x) := \sqrt{2} - \frac{1}{2^{x-\frac{1}{2}}}$ and let $\epsilon$ be the function specified by
\[
\epsilon(g(x)) = g(x) - \dfrac{1}{2}\Big(g(x - \dfrac{1}{2}) + \cdot g(x + \dfrac{1}{2}) \Big) = 
\dfrac{3 - 2\sqrt{2}}{2^{x+1}}.
\]
We define a partial antitone ranking function as follows:
\begin{align*}
\Xi \, l 
&\mapsto 
g(l)
\\
\tif{l \leq 0}{0}{\Xi \, (l - \tsample) \oplus_{1/2} \Xi \, (l + \tsample)}
&\mapsto
\frac{1}{2}\big(g(l - \frac{1}{2}) + \cdot g(l + \frac{1}{2}) \big)
\\
r \leq 0, \quad r &\mapsto 0
\end{align*}
\end{example}

>>>>

\paragraph{}
In both of these cases, there are only some values of the ranking function that are semantically important. Define a \emph{partial ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually \akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and ast, but that doesn't actually work because of terms reachable with 0 probability. Of course, astness doesn't imply rankability, so any definition of partial ranking function that includes all Y-past terms wouldn't always extend to a total ranking function.} reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
A partial ranking function that is total is just a ranking function. Providing a partial ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.

@@

\paragraph{Results to prove:}
\begin{enumerate}
\item (AST Soundness) If a closed SPCF term $M$ is antitone rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST). \lo{This should be fine.}

@@

For a more complex example, consider the following ``continuous random walk''
\[
\underbrace{\big
(\tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample) \oplus_{1/2} f(x + \tsample)} \big)}_{\Xi} \, 10
\]
Let $g(x) := \sqrt{2} - \frac{1}{2^{x-\frac{1}{2}}}$ and let $\epsilon$ be the function specified by
\[
\epsilon(g(x)) = g(x) - \dfrac{1}{2}\Big(g(x - \dfrac{1}{2}) + \cdot g(x + \dfrac{1}{2}) \Big) = 
\dfrac{3 - 2\sqrt{2}}{2^{x+1}}.
\]
We define a partial antitone ranking function as follows:
\begin{align*}
\Xi \, l 
&\mapsto 
g(l)
\\
\tif{l \leq 0}{0}{\Xi \, (l - \tsample) \oplus_{1/2} \Xi \, (l + \tsample)}
&\mapsto
\frac{1}{2}\big(g(l - \frac{1}{2}) + \cdot g(l + \frac{1}{2}) \big)
\\
r \leq 0, \quad r &\mapsto 0
\end{align*}
\end{example}

\begin{example}[Fair-in-the-limit random walk]
\label{ex:Fair-in-the-limit random walk}\citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
\[
\underbrace{\big
(\tY \, f \, x . 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{n}{2n+1}} f(x + 1)} \big)}_{\Xi} 
\, 10
\]
To construct an antitone ranking function, we solve the recurrence relation:
\[
\left\{
\begin{array}{rll}
z_0 &=& 0\\
n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
\end{array}
\right.
\]
It turns out the solution to Example $M_1$, $z_n = \frac{2^n - 1}{2^{n-1}}$, is nearly a solution in that

For $M_1$, define the function $g_1 : \nnReal \to [0, 2)$ by $g_1(x) = 2 - \frac{1}{2^{x-1}}$.
Note that $(g_1 (n))_{n \in \omega}$ satisfies the recurrence relation
Set the antitone function $\epsilon_1 (y) :=  \frac{2-y}{4}$.
Then
\begin{equation}
\epsilon_1(g_1(x)) = g_1(x) - \dfrac{1}{2} \, \big(g_1(x-1) + g_1(x+1)\big) = \dfrac{1}{2^{x+1}}
\label{eqn:epsilon g M1}
\end{equation}

\end{example}
